---
title: "ST117 Individual DRAFT Written Report - Phase 2 Task 4 (Fine grains - Butterflies)"
author: "Report Pod MS-10"
date: "08-05-2026" 
output: pdf_document
geometry: top=1cm, bottom=1cm
---


```{r setup, include=FALSE}
#Code to setup packages should go here, for example
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(lubridate)
library(broom)
library(ggfortify)
```


# a)

```{r a}
# Read Datasets
fineGrain <- read.csv("Vegetation/ECN_VF1.csv")
butterflies <- read.csv("Invertebrates/ECN_IB1.csv")

# Sort Butterflies by Year
butterflies$SDATE = as.Date(butterflies$SDATE, format = "%d-%b-%y")
butterflies$YEAR <- year(butterflies$SDATE) 

#Sum all butterfly counts per site per year across all species and transect sections
butterfly_summary <- butterflies %>%
  group_by(SITECODE, YEAR) %>%
  summarise(
    Total_Butterflies = sum(VALUE, na.rm = TRUE), 
    .groups = "drop"
  )

# Count unique species per site per year as a measure of plant diversity
fineGrain_summary <- fineGrain %>%
  filter(!is.na(VALUE)) %>%          
  group_by(SITECODE, SYEAR) %>%
  summarise(
    Unique_FG_Species = n_distinct(FIELDNAME),   
    .groups = "drop"
  ) %>%
  rename(YEAR = SYEAR)

# Constructed Overall Dataset
combined_dataset <- butterfly_summary %>%
  left_join(fineGrain_summary, by = c("SITECODE", "YEAR"))


kable(head(combined_dataset, 3), caption = "First Three Rows of Combined Dataset")


```
\newpage
# b)

````{r b, warning=FALSE, message=FALSE}
# Plot
ggplot(combined_dataset, aes(x = Unique_FG_Species, y = Total_Butterflies, 
                              color = SITECODE)) +
  geom_point() +
  labs(
    title = "Fine Grain Species Richness vs Butterfly Abundance",
    x = "Number of Unique Fine Grain Species",
    y = "Total Butterfly Count"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

````
The scatter plot reveals a weak positive relationship between fine-grain species 
richness and butterfly abundance. Sites with low species richness consistently record 
low butterfly counts, whilst at higher richness values the butterfly counts become 
considerably more variable, ranging from very low to extremely high. 
Colouring points by SITECODE reveals that certain sites such as T05 and T10 
consistently record higher butterfly counts regardless of fine grain species richness, 
suggesting site-level differences may be a more important driver of butterfly 
abundance than plant diversity alone.

\newpage
# c)

````{r c, warning=FALSE, message=FALSE}

ggplot(combined_dataset, aes(x = Unique_FG_Species, y = Total_Butterflies)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Fine Grain Species Richness vs Butterfly Abundance",
    x = "Number of Unique Fine Grain Species",
    y = "Total Butterfly Count"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

# Remove the outlier species

# Finding Quartiles
Q1 <- quantile(combined_dataset$Total_Butterflies, 0.25, na.rm = TRUE)
Q3 <- quantile(combined_dataset$Total_Butterflies, 0.75, na.rm = TRUE)

IQR_value <- IQR(combined_dataset$Total_Butterflies, na.rm = TRUE)

# Bounds for Outliers
lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

# Removing Outliers
data_no_outlier <- combined_dataset %>%
  filter(
    Total_Butterflies >= lower_bound &
    Total_Butterflies <= upper_bound
  )

# Plot Without Outliers
ggplot(data_no_outlier, aes(x = Unique_FG_Species, y = Total_Butterflies)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Relationship Without Extreme Outlier",
    x = "Number of Unique Fine Grain Species",
    y = "Total Butterfly Count"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
# Models (with and without outliers)
model <- lm(Total_Butterflies ~ Unique_FG_Species, data = combined_dataset)
model_no_outliers <- lm(Total_Butterflies ~ Unique_FG_Species, data = data_no_outlier)
r_squared <- summary(model)$r.squared
r_squared_no_outliers <- summary(model_no_outliers)$r.squared

# Further Info Plots on Overall Model
autoplot(model, which = 1:2)

# Histogram - Showing skewness and presence of extreme values
ggplot(combined_dataset, aes(x = Total_Butterflies)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 10) +
  labs(title = "Distribution of Total Butterfly Counts",
       x = "Total Butterfly Count", y = "Frequency") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

# Table showing Gradient and Intercept of the Regression lines
table_model <- bind_rows(
  tidy(model) %>% mutate(Model = "With Outlier"),
  tidy(model_no_outliers) %>% mutate(Model = "Without Outlier")
)

table_model <- table_model %>%
  mutate(term = case_when(
    term == "(Intercept)" ~ "Intercept",
    term == "Unique_FG_Species" ~ "Gradient",
    TRUE ~ term
  ))

kable(table_model, digits = 3, caption = "Linear Model Coefficients With and Without Outliers")


````
The initial model suggests a slight negative relationship between fine-grain species 
richness and butterfly abundance, though this is not statistically significant 
(gradient = -40.35, p = 0.489). Interestingly, once outliers are removed the gradient 
actually flips direction (gradient = 22.47, p = 0.535), remaining non-significant. 
This instability in both the magnitude and direction of the gradient is concerning and 
suggests the apparent relationship is highly sensitive to a small number of extreme 
observations rather than reflecting a genuine underlying trend in the data.

The diagnostic plots raise further concerns about the validity of the model. The Q-Q 
plot shows that residuals deviate substantially from normality, particularly in the 
upper tail, whilst the residuals vs fitted plot reveals clear heteroscedasticity, with 
variance increasing at higher fitted values. The histogram of total butterfly counts 
confirms the response variable is heavily right-skewed, which partly explains these 
violations. Together with a low R² of `r round(r_squared, 3)`, these diagnostics suggest that simple linear 
regression is not well-suited to this data. Butterfly abundance is likely influenced 
by more complex ecological factors such as site-level characteristics, weather 
conditions, or habitat quality that a single predictor model cannot adequately capture.

\newpage
# d)


````{r d, warning=FALSE, message=FALSE, results='asis'}
# Remove rows with missing fine-grain data before fitting site models
data_fg <- combined_dataset %>%
  filter(!is.na(Unique_FG_Species))

sites_fg <- unique(data_fg$SITECODE)

for (site in sites_fg) {
  
  site_data <- data_fg %>% filter(SITECODE == site)
  
  # Skip if not enough data
  if (nrow(site_data) < 2) {
    cat(paste("\nSite", site, "- not enough data, so skipping\n"))
    next
  }
  
  model_site <- lm(Total_Butterflies ~ Unique_FG_Species, data = site_data)
  
  # Scatter + regression line per site
  print(
    ggplot(site_data, aes(x = Unique_FG_Species, y = Total_Butterflies)) +
      geom_point(color = "blue") + geom_smooth(method = "lm", color = "red", se = FALSE) +
      labs(title = paste("Site", site, ": Fine Grain Species Richness vs Butterfly Abundance"),
        x = "Number of Unique Fine Grain Species",
        y = "Total Butterfly Count"
      ) + theme_classic() + theme(plot.title = element_text(hjust = 0.5))
  )
  
  # Coefficient table per site
  print(
    kable(tidy(model_site), digits = 3,
          caption = paste("Linear Model Coefficients for Site", site))
  )
}
````
Site-specific models reveal little to no consistent relationship between fine-grain 
species richness and butterfly abundance across the network. Most slopes are non-significant. Site T06 had no variation in fine-grain species richness (all observations recorded exactly one species), making the slope unidentifiable. T12 had only two observations, producing a perfect fit with undefined standard errors and no meaningful inference possible. 
The site-specific regression lines deviate considerably from the global model in 
several panels, confirming the global model poorly represents individual sites. 
This is consistent with part c), where the apparent relationship appears driven by 
outliers and heterogeneity between sites rather than a genuine network-wide trend.